workout_probabilities

### 2019-04-24 workout_probabilities.Rmd
still going

The likelihood under the hypothesis is then $\mathcal{L}_{\nu,m}(\mathbf{x}) = \prod_{i=1}^{D} \textrm{f}_{\nu,m}(x_i)$.
MuTect reports the log likelihood ratio $log(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or t_lod_fstar depending on the version.
By fixing the threshold posterior odds at two, the site-specific mutation probability a constant $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$, and $\mathrm{p}(m \mid M)$ the prior probability of mutation to specific allele $m$ constant $\mathrm{p}(m \mid M) = \mu/3 = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3 for classifying a site as a somatic variant.
Here we examine the effect of the assumption of a constant prior probability of mutation.

@@

The posterior distribution of $\boldsymbol{\pi}$ is $\boldsymbol{\pi} \mid C,\boldsymbol{\alpha} \sim \textrm{Dirichlet}(\mathbf{C} + \boldsymbol{\alpha})$, where $\mathbf{C} = (C_1,\dots,C_{96})$ are the counts of mutations present in the tumor for each of the 96 contexts.

>>>>

The likelihood under the hypothesis is then $\mathcal{L}_{\nu,m}(\mathbf{x}) = \prod_{i=1}^{D} \textrm{f}_{\nu,m}(x_i)$.
MuTect reports the log likelihood ratio $\mathrm{log}(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or t_lod_fstar depending on the version.
By fixing the threshold posterior odds at two, the site-specific mutation probability a constant $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$, and $\mathrm{p}(m \mid M)$ the prior probability of mutation to specific allele $m$ constant $\mathrm{p}(m \mid M) = \mu/3 = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3 for classifying a site as a somatic variant.
Here we examine the effect of the assumption of a constant prior probability of mutation.

@@

The posterior distribution of $\boldsymbol{\pi}$ is $\boldsymbol{\pi} \mid C,\boldsymbol{\alpha} \sim \textrm{Dirichlet}(\mathbf{C} + \boldsymbol{\alpha})$, where $\mathbf{C} = (C_1,\dots,C_{96})$ are the counts of mutations present in the tumor for each of the 96 contexts.
We compute an empirical bayes estimate of $\boldsymbol{\pi}$ by choosing $\mathbf{C}$ as the set of mutations assigned a TLOD by MuTect above some threshold, which we choose as 10.
We show through extensive simulation that our estimate of $\boldsymbol{\pi}$ converges quickly its true simulated value.

Returning to the model above, we can calculate the log posterior odds in favor of $\mathbf{H_1}$ as

\[
  \textrm{log}_{10} \left(\frac{(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})\mathrm{p}(m,M \mid C)}{(\mathcal{L}_{\nu=0,m}(\mathbf{x})(1-\mathrm{p}(m,M \mid C))} \right) = \textrm{TLOD} + \textrm{log prior odds},
\]

and the posterior odds ratio in favor of $\mathbf{H_1}$ as
\[
10^{(\textrm{TLOD} + \textrm{log prior odds})}.
\]

We show via extensive simulation that under the assumption that the mutation signature describes the biological process generating mutations in a tumor, our posterior odds ratio is a better classifier than MuTect at any threshold.
